A dynamic bioenergetic model for coral-*Symbiodinium* symbioses and coral bleaching as an alternate stable state

### 2017-03-21 JTB_ms/JTB_ms.Rmd
update fig caps, yC effects

---
title: A dynamic bioenergetic model for coral-*Symbiodinium* symbioses and coral bleaching as an alternate stable state
author:
  - name: Ross Cunning
    email: ross.cunning@gmail.com
    affiliation: University of Hawaii
    footnote: Corresponding Author
  - name: Erik B. Muller
    affiliation: UCSB, MSI
  - name: Ruth D. Gates
    affiliation: HIMB
  - name: Roger M. Nisbet
    affiliation: UCSB
address:
  - code: HIMB
    address: Hawaii Institute of Marine Biology, School of Ocean and Earth Science and Technology, University of Hawaii at Manoa, USA
  - code: UCSB
    address: Department of Ecology, Evolution, and Marine Biology, University of California, Santa Barbara, USA
  - code: MSI
    address: Marine Science Institute, University of California, Santa Barbara, USA
abstract: |
  Coral reef ecosystems owe their ecological success--and vulnerability to climate change--to the symbiotic metabolism of corals and *Symbiodinium* spp. The urgency to understand and predict the stability and breakdown of these symbioses (i.e., coral 'bleaching') demands the development and application of theoretical tools. Here, we develop a dynamic bioenergetic model of coral-*Symbiodinium* symbioses that demonstrates realistic steady-state patterns in coral growth and symbiont abundance across gradients of light, nutrients, and feeding. Furthermore, by including a mechanistic treatment of photo-oxidative stress, the model displays dynamics of bleaching and recovery that can be explained as transitions between alternate stable states. These dynamics reveal that 'healthy' and 'bleached' states correspond broadly to nitrogen- and carbon-limitation of the system, with transitions between them occurring as integrated responses to multiple environmental factors. Indeed, a suite of complex emergent behaviors reproduced by the model (e.g., bleaching is exacerbated by nutrients and attenuated by feeding) suggests it unifyingly captures many important attributes of the system; meanwhile, its modular framework and open source R code are designed to facilitate further problem-specific development. We see significant potential for this modeling framework to generate testable hypotheses and predict integrated, mechanistic responses of corals to environmental change, with important implications for understanding the performance and maintenance of symbiotic systems.
bibliography: library.bib
output: 
  rticles::elsevier_article:
    fig_caption: yes
---
```{r setup, include=FALSE}
library(reshape2); library(dplyr)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

@@

| Symbol  | Description                          | Units                 | Eq. no. |
| ------- | ------------------------------------ | --------------------- | ------- |
| $j_X$     | Prey uptake rate                       | mol X C-mol H^-1^ d^-1^ | 3 |
| $j_N$     | Nitrogen uptake rate                   | mol N Cmol H^-1^ d^-1^ | 4 |
| $j_{HG}$  | Host biomass formation rate            | C-mol H C-mol H^-1^ d^-1^| 5 |
| $j_{HT}$  | Host biomass turnover rate             | C-mol H C-mol H^-1^ d^-1^| 6 |
| $r_{NH}$  | Recycled nitrogen from host turnover   | mol N C-mol H^-1^ d^-1^ | 7 |
| $\rho_N$  | Nitrogen shared with the symbiont      | mol N C-mol H^-1^ d^-1^ | 8 |
| $j_{eC}$  | Excess carbon used to activate host CCMs| mol C C-mol H^-1^ d^-1^ | 9 |
| $j_{CO_2}$| CO~2~ input to photosynthesis          | mol CO~2~ C-mol H^-1^ d^-1^| 10 |
| $j_L$     | Light absorption rate                  | mol photons C-mol S^-1^ d^-1^| 12 |
| $r_{CH}$  | Recycled CO~2~ from host turnover      | mol CO~2~ C-mol H^-1^ d^-1^| 13 |
| $r_{CS}$  | Recycled CO~2~ from symbiont turnover  | mol CO~2~ C-mol S^-1^ d^-1^| 14 |
| $j_{CP}$  | Photosynthesis rate                    | mol C C-mol S^-1^ d^-1^ | 15 |
| $j_{eL}$  | Light energy in excess of photochemistry| mol photons C-mol S^-1^ d^-1^| 16 |
| $j_{NPQ}$ | Total capacity of NPQ                  | mol photons C-mol S^-1^ d^-1^ | 17 |
| $c_{ROS}$ | ROS production proportional to baseline | -- | 18 |
| $r_{NS}$  | Recycled nitrogen from symbiont turnover| mol N C-mol S^-1^ d^-1^| 19 |
| $j_{SG}$  | Symbiont biomass formation rate        | C-mol S C-mol S^-1^ d^-1^| 20 |
| $\rho_C$  | Fixed carbon shared with host          | mol C C-mol S^-1^ d^-1^ | 21 |
| $j_{ST}$  | Symbiont biomass turnover rate         | C-mol S C-mol S^-1^ d^-1^| 22 |

@@

| Symbol  | Description                          | Units               | Value     |
| ------- | ------------------------------------ | ------------------- | --------- |
| $n_{NH}$ | N:C molar ratio in host biomass | -- | $`r defpars$nNH`$ |
| $n_{NS}$ | N:C molar ratio in symbiont biomass | -- | $`r defpars$nNS`$ |
| $n_{NX}$ | N:C molar ratio in prey biomass | -- | $`r defpars$nNX`$ |
| $j_{HT}^0$ | Specific maintenance rate of host biomass | C-mol H C-mol H^-1^ d^-1^ | $`r defpars$jHT0`$ |
| $j_{ST}^0$ | Specific maintenance rate of symbiont biomass | C-mol S C-mol S^-1^ d^-1^ | $`r defpars$jST0`$ |
| $\sigma_{NH}$ | Proportion host N-turnover recycled to host | -- | $`r defpars$sigmaNH`$ |
| $\sigma_{CH}$ | Proportion host C-turnover recycled to photosynthesis | -- | $`r defpars$sigmaCH`$ |
| $\sigma_{NS}$ | Proportion symbiont N-turnover recycled to symbiont | -- | $`r defpars$sigmaNS`$ |
| $\sigma_{CS}$ | Proportion symbiont C-turnover recycled to photosynthesis | -- | $`r defpars$sigmaCS`$ |
| $j_{Xm}$ | Maximum specific feeding rate of host | molX C-mol H^-1^ d^-1^ | $`r defpars$jXm`$ |
| $K_X$ | Half-saturation constant for prey uptake by host | molX L^-1^ | $`r defpars$KX`$ |
| $j_{Nm}$ | Maximum specific DIN uptake rate by host | mol N C-mol H^-1^ d^-1^ | $`r defpars$jNm`$ |
| $K_N$ | Half-saturation constant for DIN uptake by host | mol N L^-1^ | $`r defpars$KN`$ |
| $k_{CO_2}$ | Efficacy of CO~2~ delivery by host CCMs | mol CO~2~ mol C^-1^ | $`r defpars$kCO2`$ |
| $j_{HGm}$ | Maximum specific growth rate of host | C-mol H C-mol H^-1^ d^-1^ | $`r defpars$jHGm`$ |
| $y_{CL}$ | Quantum yield of photosynthesis | mol C mol photons^-1^ | $`r defpars$yCL`$ |
| $\bar{a}^*$ | Effective light-absorbing cross-section of symbiont | m^2^ C-mol S^-1^ | $`r defpars$astar`$ |
| $k_{NPQ}$ | NPQ capacity of symbiont | mol photons C-mol S^-1^ d^-1^ | $`r defpars$kNPQ`$ |
| $k_{ROS}$ | Excess photon energy that doubles ROS production, relative to baseline levels | mol photons C-mol S^-1^ d^-1^ | $`r defpars$kROS`$ |
| $j_{CPm}$ | Maximum specific photosynthesis rate of symbiont | mol C C-mol S^-1^ d^-1^ | $`r defpars$jCPm`$ |
| $j_{SGm}$ | Maximum specific growth rate of symbiont | C-mol S C-mol S^-1^ d^-1^ | $`r defpars$jSGm`$ |
| $b$ | Scaling parameter for bleaching response | -- | $`r defpars$b`$ |

![Graphical representation of coral-algal symbiosis model. Light, CO~2~, prey, and DIN are acquired from the external environment proportional to the biomass of the partner indicated by the black box for uptake. Mass fluxes (see Table 1 for definitions) are represented by $j$'s with subscripts indicating the type of mass, and in some cases the process (e.g., $j_{CP}$ is the flux of carbon produced by photosynthesis), and $\rho$'s indicate fluxes that are shared by one partner with the other. Parallel complementary synthesizing units (SUs) are represented by large circles, and rejection fluxes from these SUs are indicated by dashed lines. $c_{ROS}$ is a proportional rate that impacts other model fluxes by inhibition or acceleration; likewise, $j_{eC}$ accelerates the rate of $j_{CO_2}$. Recycled mass fluxes from biomass turnover are not shown for clarity (but see Table 1 for definitions).](../img/Fig1.png)

@@

The coral animal acquires both carbon and nitrogen from feeding on prey from the environment. Prey acquisition is specified by Michaelis-Menten kinetics (i.e., a Holling type II function) with a maximum feeding rate $j_{Xm}$ and half-saturation constant $K_X$:

@@

\begin{equation} j_{HG} = \bigg({1 \over j_{HGm}} + {1 \over {\rho_C{S \over H} + j_X}} + {1 \over {(j_N + n_{NX}j_X + r_{NH}) / n_{NH}}} - {1 \over {{\rho_C{S \over H} + j_X} + (j_N + n_{NX}j_X + r_{NH}) / n_{NH}}}\bigg)^{-1} \end{equation}

where $\rho_C$ is fixed carbon shared by the symbiont (see Eq. 21), and $r_{NH}$ is recycled nitrogen liberated by host biomass turnover (see Eq. 7). Host biomass turnover is equal to the specific maintenance rate of host biomass,

@@

\begin{equation} \rho_N = (j_N + n_{NX}j_X + r_{NH} - n_{NH}j_{HG})_+ \end{equation}

@@

\begin{equation} j_{eC} = (jX + \rho_C{S \over H} - j_{HG})_+ \end{equation}

@@

CO~2~ arrives at the photosynthesis SU from multiple sources: in addition to the CO~2~ actively supplied by the host through its CCMs ($j_{CO_2}$; Eq. 10), we assume a fixed proportion $\sigma_{CH}$ of metabolic CO~2~ generated by host biomass turnover is passively available to the photosynthesis SU, according to:

\begin{equation} r_{CH}=\sigma_{CH}j_{HT} \end{equation}

along with a fixed proportion of CO~2~ generated by symbiont biomass turnover[^2]:

\begin{equation} r_{CS}=\sigma_{CS}j_{ST}^0 \end{equation}

@@

where $k_{ROS}$ is a parameter of the symbiont that determines the rate of ROS production (specifically, the amount of excess excitation energy that doubles ROS production relative to baseline levels). Importantly, $c_{ROS}$ is specified here not as a function of absolute external light, but rather the amount of excess light energy after accounting for quenching by carbon fixation and NPQ. A direct consequence of this formulation is that CO~2~-limitation of photosynthesis can lead to ROS production, an important mechanism [@Wooldridge:2009p7807] that was not captured by previous representations of photooxidative stress [@Eynaud:2011tv]. With this single SU, both the light and dark reactions of photosynthesis are represented, allowing for sink-limitation (i.e., CO~2~-limitation) to cause overreduction of the electron transport chain and ROS production.

@@

\begin{equation} j_{SG} = \bigg({1 \over j_{SGm}} + {1 \over j_{CP}} + {1 \over {(\rho_N{H \over S} + r_{NS}) / n_{NS}}} - {1 \over {j_{CP} + (\rho_N{H \over S} + r_{NS}) / n_{NS}}}\bigg)^{-1} \end{equation}

@@

\begin{equation} \rho_C = (j_{CP} - j_{SG})_+ \end{equation}

@@

where $j_{S1}$ and $j_{S2}$ are the specific input fluxes of the two substrates and $j_{Pm}$ is the maximum specific product formation rate, in units of Cmol Cmol^-1^ d^-1^.

@@

![Alternate stable states in S:H biomass and growth across a light gradient. Alternate stable states occur between ~33-44 mol photons m^-2^ d^-1^ under these conditions (DIN=1e-7 molN L^-1^; X=2e-6 molX L^-1^), depending on whether initial S:H is high (0.1, closed circles), representing a healthy coral, or low (0.0001, open circles), representing a bleached coral.](../img/Fig2.png)

@@

![Steady state values of (A) specific growth (Cmol Cmol^-1^ d^-1^) and (B) the symbiont to host biomass ratio (C-mol S C-mol H^-1^) across gradients of external irradiance and dissolved inorganic nitrogen. Note that typical conditions for reefs are ~1e-7 M DIN and 10-20 mol photons m^-2^ d^-1^. Simulations for each combination of light and nutrients (41 points along each axis) were run for 100 days with a time step of 1 day. All parameters were set to default values and prey density set to zero. Negative steady state growth rates, and corresponding S:H ratios, were set to zero.](../img/Fig3.png)

@@

![Sensitivity analysis. Plots show the fractional change in steady state values of growth (solid lines) and S:H biomass (dashed lines) in response to fractional changes in default parameter values (see Table 2 for default values). Parameters are grouped by which processes they are involved in. This sensitivity analysis was conducted at conditions typical for coral reef environments: low DIN (1e-7 M) and intermediate light (15 mol m^-2^ d^-2^). Sensitivity analyses conducted other environmental conditions are presented in Figs S3-S7.](../img/Fig4.png)

@@

![Light-driven seasonal dynamics of symbiont abundance and coral growth. Light input (upper panel) was designed as a sinusoidal curve with a period of one year, with maximum and minimum values of 44 and 22 mol photons m^-2^ d^-1^, based on @Stimson:1997p3837. The dynamic behavior of symbiont to host biomass ratio (middle panel) and the specific growth rate of host biomass (lower panel) show seasonal oscillations that are greater in magnitude under higher nutrients (1e-5 mol N L^-1^; dashed lines) relative to lower nutrients (1e-7 mol N L^-1^; solid lines), consistent with the findings of @Stimson:1997p3837. Prey density was set at 1e-5 CmolX L^-1^.](../img/Fig5.png)

@@

![Coral bleaching as a switch from a nitrogen- to carbon-limited alternative stable state. This transition is demonstrated in response to gradually increasing external light (A), which causes an increase in production of ROS (B) that reduces the photosynthetic rate through photoinhibition (C). Decreasing photosynthesis moves the system from overall nitrogen-limitation toward carbon-limitation (D); when this threshold is crossed, the system rapidly becomes highly carbon-limited as photosynthesis becomes CO~2~-limited (E) and symbiont densities rapidly decline (F) into a bleached state. Substrate limitation coefficients were calculated using Eq. 23. All parameters were set at default values with external DIN=1e-7 mol N/L and prey density=0).](../img/Fig6.png)

@@

![Bleaching with interactive factors. Simulations of high light stress (sinusoid with maximum=48 mol ph m^-2^ d^-1^) under default environmental conditions (solid line; DIN=1e-7 mol N L^-1^; prey=3e-6 CmolX L^-1^), or with elevated feeding (dashed line; prey=6e-6 CmolX L^-1^), or elevated nutrients (dotted line; DIN=1e-6 mol N L^-1^). Initial symbiont biomass was set to the steady state for each set of starting conditions, respectively. (All parameters set to default values).](../img/Fig7.png)

@@

![Recovery from bleaching with varying N:C availability (A: N:C=0.09; B: N:C=0.10; C: N:C=0.30). Higher N:C ratios in a heterotrophic food source (effectively representing higher external DIN and/or lower feeding rates) cause a larger overshoot in the S:H biomass ratio, and prolong the duration of time until the system 'recovers' by re-establishing nitrogen-limitation. Substrate limitation coefficients were calculated using Eq. 23. All simulations were run with default parameters (except varying $n_{NX}$), with L=10 mol photons m^2^ d^-1^, DIN=1e-7 mol N L^-1^, prey=1e-6 CmolX L^-1^, and initial S:H biomass=0.001.](../img/Fig8.png)

>>>>

---
title: A dynamic bioenergetic model for coral-*Symbiodinium* symbioses and coral bleaching as an alternate stable state
author:
  - name: Ross Cunning
    email: ross.cunning@gmail.com
    affiliation: University of Hawaii
    footnote: Corresponding Author
  - name: Erik B. Muller
    affiliation: UCSB, MSI
  - name: Ruth D. Gates
    affiliation: HIMB
  - name: Roger M. Nisbet
    affiliation: UCSB
address:
  - code: HIMB
    address: Hawaii Institute of Marine Biology, School of Ocean and Earth Science and Technology, University of Hawaii at Manoa, USA
  - code: UCSB
    address: Department of Ecology, Evolution, and Marine Biology, University of California, Santa Barbara, USA
  - code: MSI
    address: Marine Science Institute, University of California, Santa Barbara, USA
abstract: |
  Coral reef ecosystems owe their ecological success--and vulnerability to climate change--to the symbiotic metabolism of corals and *Symbiodinium* spp. The urgency to understand and predict the stability and breakdown of these symbioses (i.e., coral 'bleaching') demands the development and application of theoretical tools. Here, we develop a dynamic bioenergetic model of coral-*Symbiodinium* symbioses that demonstrates realistic steady-state patterns in coral growth and symbiont abundance across gradients of light, nutrients, and feeding. Furthermore, by including a mechanistic treatment of photo-oxidative stress, the model displays dynamics of bleaching and recovery that can be explained as transitions between alternate stable states. These dynamics reveal that 'healthy' and 'bleached' states correspond broadly to nitrogen- and carbon-limitation in the system, with transitions between them occurring as integrated responses to multiple environmental factors. Indeed, a suite of complex emergent behaviors reproduced by the model (e.g., bleaching is exacerbated by nutrients and attenuated by feeding) suggests it captures many important attributes of the system; meanwhile, its modular framework and open source R code are designed to facilitate further problem-specific development. We see significant potential for this modeling framework to generate testable hypotheses and predict integrated, mechanistic responses of corals to environmental change, with important implications for understanding the performance and maintenance of symbiotic systems.
bibliography: library.bib
output: 
  rticles::elsevier_article:
    fig_caption: yes
---
```{r setup, include=FALSE}
library(reshape2); library(dplyr)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

@@

| Symbol  | Description                          | Units                 | Eq. no. |
| ------- | ------------------------------------ | --------------------- | ------- |
| $j_X$     | Prey uptake rate                       | mol X C-mol H^-1^ d^-1^ | 3 |
| $j_N$     | Nitrogen uptake rate                   | mol N Cmol H^-1^ d^-1^ | 4 |
| $j_{HG}$  | Host biomass formation rate            | C-mol H C-mol H^-1^ d^-1^| 5 |
| $j_{HT}$  | Host biomass turnover rate             | C-mol H C-mol H^-1^ d^-1^| 6 |
| $r_{NH}$  | Recycled nitrogen from host turnover   | mol N C-mol H^-1^ d^-1^ | 7 |
| $\rho_N$  | Nitrogen shared with the symbiont      | mol N C-mol H^-1^ d^-1^ | 8 |
| $j_{eC}$  | Excess carbon used to activate host CCMs| mol C C-mol H^-1^ d^-1^ | 9 |
| $j_{CO_2}$| CO~2~ input to photosynthesis          | mol CO~2~ C-mol H^-1^ d^-1^| 10 |
| $j_L$     | Light absorption rate                  | mol photons C-mol S^-1^ d^-1^| 12 |
| $r_{CH}$  | Recycled CO~2~ from host               | mol CO~2~ C-mol H^-1^ d^-1^| 13 |
| $r_{CS}$  | Recycled CO~2~ from symbiont           | mol CO~2~ C-mol S^-1^ d^-1^| 14 |
| $j_{CP}$  | Photosynthesis rate                    | mol C C-mol S^-1^ d^-1^ | 15 |
| $j_{eL}$  | Light energy in excess of photochemistry| mol photons C-mol S^-1^ d^-1^| 16 |
| $j_{NPQ}$ | Total capacity of NPQ                  | mol photons C-mol S^-1^ d^-1^ | 17 |
| $c_{ROS}$ | ROS production proportional to baseline | -- | 18 |
| $r_{NS}$  | Recycled nitrogen from symbiont turnover| mol N C-mol S^-1^ d^-1^| 19 |
| $j_{SG}$  | Symbiont biomass formation rate        | C-mol S C-mol S^-1^ d^-1^| 20 |
| $\rho_C$  | Fixed carbon shared with host          | mol C C-mol S^-1^ d^-1^ | 21 |
| $j_{ST}$  | Symbiont biomass turnover rate         | C-mol S C-mol S^-1^ d^-1^| 22 |

@@

| Symbol  | Description                          | Units               | Value     |
| ------- | ------------------------------------ | ------------------- | --------- |
| $n_{NH}$ | N:C molar ratio in host biomass | -- | $`r defpars$nNH`$ |
| $n_{NS}$ | N:C molar ratio in symbiont biomass | -- | $`r defpars$nNS`$ |
| $n_{NX}$ | N:C molar ratio in prey biomass | -- | $`r defpars$nNX`$ |
| $j_{HT}^0$ | Specific maintenance rate of host biomass | C-mol H C-mol H^-1^ d^-1^ | $`r defpars$jHT0`$ |
| $j_{ST}^0$ | Specific maintenance rate of symbiont biomass | C-mol S C-mol S^-1^ d^-1^ | $`r defpars$jST0`$ |
| $\sigma_{NH}$ | Proportion host N-turnover recycled to host | -- | $`r defpars$sigmaNH`$ |
| $\sigma_{CH}$ | Proportion host C-turnover recycled to photosynthesis | -- | $`r defpars$sigmaCH`$ |
| $\sigma_{NS}$ | Proportion symbiont N-turnover recycled to symbiont | -- | $`r defpars$sigmaNS`$ |
| $\sigma_{CS}$ | Proportion symbiont C-turnover recycled to photosynthesis | -- | $`r defpars$sigmaCS`$ |
| $j_{Xm}$ | Maximum specific assimilation rate from host feeding | molX C-mol H^-1^ d^-1^ | $`r defpars$jXm`$ |
| $K_X$ | Half-saturation constant for prey uptake by host | molX L^-1^ | $`r defpars$KX`$ |
| $j_{Nm}$ | Maximum specific DIN uptake rate by host | mol N C-mol H^-1^ d^-1^ | $`r defpars$jNm`$ |
| $K_N$ | Half-saturation constant for DIN uptake by host | mol N L^-1^ | $`r defpars$KN`$ |
| $k_{CO_2}$ | Efficacy of CO~2~ delivery by host CCMs | mol CO~2~ mol C^-1^ | $`r defpars$kCO2`$ |
| $j_{HGm}$ | Maximum specific growth rate of host | C-mol H C-mol H^-1^ d^-1^ | $`r defpars$jHGm`$ |
| $y_{CL}$ | Quantum yield of photosynthesis | mol C mol photons^-1^ | $`r defpars$yCL`$ |
| $y_{C}$ | Yield of biomass formation from carbon | C-mol mol C^-1^ | $`r defpars$yC`$ |
| $\bar{a}^*$ | Effective light-absorbing cross-section of symbiont | m^2^ C-mol S^-1^ | $`r defpars$astar`$ |
| $k_{NPQ}$ | NPQ capacity of symbiont | mol photons C-mol S^-1^ d^-1^ | $`r defpars$kNPQ`$ |
| $k_{ROS}$ | Excess photon energy that doubles ROS production, relative to baseline levels | mol photons C-mol S^-1^ d^-1^ | $`r defpars$kROS`$ |
| $j_{CPm}$ | Maximum specific photosynthesis rate of symbiont | mol C C-mol S^-1^ d^-1^ | $`r defpars$jCPm`$ |
| $j_{SGm}$ | Maximum specific growth rate of symbiont | C-mol S C-mol S^-1^ d^-1^ | $`r defpars$jSGm`$ |
| $b$ | Scaling parameter for bleaching response | -- | $`r defpars$b`$ |

![Graphical representation of coral-algal symbiosis model. Light, CO~2~, prey, and DIN are acquired from the external environment proportional to the biomass of the partner indicated by the black box for uptake. Mass fluxes (see Table 1 for definitions) are represented by $j$'s with subscripts indicating the type of mass, and in some cases the process (e.g., $j_{CP}$ is the flux of carbon produced by photosynthesis), and $\rho$'s indicate fluxes that are shared by one partner with the other. Parallel complementary synthesizing units (SUs) are represented by large circles, and rejection fluxes from these SUs are indicated by dashed lines. $c_{ROS}$ is a proportional rate that impacts other model fluxes by inhibition or acceleration; likewise, $j_{eC}$ accelerates the rate of $j_{CO_2}$. Recycling fluxes are not shown for clarity (but see Table 1 for definitions).](../img/Fig1.png)

@@

The coral animal acquires both carbon and nitrogen from feeding on prey from the environment. Assimilation from feeding is specified by Michaelis-Menten kinetics (i.e., a Holling type II function) with a maximum rate of $j_{Xm}$ and half-saturation constant $K_X$:

@@

\begin{equation} j_{HG} = \bigg({1 \over j_{HGm}} + {1 \over {y_{C}\rho_C{S \over H} + j_X}} + {1 \over {(j_N + n_{NX}j_X + r_{NH}) / n_{NH}}} - {1 \over {y_{C}{\rho_C{S \over H} + j_X} + (j_N + n_{NX}j_X + r_{NH}) / n_{NH}}}\bigg)^{-1} \end{equation}

where $\rho_C$ is fixed carbon shared by the symbiont (see Eq. 21), and $r_{NH}$ is recycled nitrogen liberated by host biomass turnover (see Eq. 7). The parameter $y_C$ specifies the yield of biomass from organic carbon, which we take to be 0.8 for redox balance (see @Muller:2009io). Host biomass turnover is equal to the specific maintenance rate of host biomass,

@@

\begin{equation} \rho_N = (j_N + n_{NX}j_X + r_{NH} - n_{NH}j_{HG}y_{C}^{-1})_+ \end{equation}

@@

\begin{equation} j_{eC} = (j_X + \rho_C{S \over H} - j_{HG}y_{C}^{-1})_+ \end{equation}

@@

CO~2~ arrives at the photosynthesis SU from multiple sources: in addition to the CO~2~ actively supplied by the host through its CCMs ($j_{CO_2}$; Eq. 10), we assume a fixed proportion $\sigma_{CH}$ of metabolic CO~2~ generated by the host from both biomass turnover and formation is passively available to the photosynthesis SU, according to:

\begin{equation} r_{CH}=\sigma_{CH}(j_{HT} + (1-y_C)j_{HG}/y_C) \end{equation}

along with a fixed proportion of CO~2~ generated by symbiont biomass turnover[^2] and formation:

\begin{equation} r_{CS}=\sigma_{CS}(j_{ST}^0 + (1-y_C)j_{SG}/y_C) \end{equation}

@@

where $k_{ROS}$ is a parameter of the symbiont that determines the rate of ROS production (specifically, the amount of excess excitation energy that doubles ROS production relative to baseline levels). Importantly, $c_{ROS}$ is specified here not as a function of absolute external light, but rather the amount of excess light energy after accounting for quenching by carbon fixation and NPQ. A direct consequence of this formulation is that CO~2~-limitation of photosynthesis can lead to ROS production, an important mechanism [@Butow:1998gz; @Wooldridge:2009p7807] that was not captured by previous representations of photooxidative stress [@Eynaud:2011tv]. With this single SU, both the light and dark reactions of photosynthesis are represented, allowing for sink-limitation (i.e., CO~2~-limitation) to cause overreduction of the electron transport chain and ROS production.

@@

\begin{equation} j_{SG} = \bigg({1 \over j_{SGm}} + {1 \over y_{C}j_{CP}} + {1 \over {(\rho_N{H \over S} + r_{NS}) / n_{NS}}} - {1 \over {y_{C}j_{CP} + (\rho_N{H \over S} + r_{NS}) / n_{NS}}}\bigg)^{-1} \end{equation}

@@

\begin{equation} \rho_C = (j_{CP} - j_{SG}y_{C}^{-1})_+ \end{equation}

@@

where $j_{S1}$ and $j_{S2}$ are the specific input fluxes of the two substrates and $j_{Pm}$ is the maximum specific product formation rate, in units of Cmol Cmol^-1^ d^-1^. When both substrate input fluxes are higher than what can be used at the maximum production rate, this limitation coefficient is zero, implying that neither substrate is limiting production.

@@

![Alternate stable states in S:H biomass and growth across a light gradient. Alternate stable states occur between ~25-42 mol photons m^-2^ d^-1^ under these conditions (DIN=1e-7 molN L^-1^; X=1e-7 CmolX L^-1^), depending on whether initial S:H is high (1, closed circles), representing a healthy coral, or low (0.0001, open circles), representing a bleached coral. Arrow above point at L=25 indicates a S:H ratio beyond the axis range; this 'overshoot' phenomenon, in which initially  bleached corals may achieve high S:H ratios while remaining in a carbon-limited state is discussed in the *Coral Bleaching and Recovery* section.](../img/Fig2.png)

@@

![Steady state values of (A) specific growth (C-mol H C-mol H^-1^ d^-1^) and (B) the symbiont to host biomass ratio (C-mol S C-mol H^-1^) across gradients of external irradiance and dissolved inorganic nitrogen. Note that typical conditions for reefs are ~1e-7 M DIN and 10-20 mol photons m^-2^ d^-1^. Simulations for each combination of light and nutrients (41 points along each axis) were run to steady state with all parameters at default values and prey density set to zero. Negative steady state growth rates, and corresponding S:H ratios, were set to zero.](../img/Fig3.png)

@@

![Sensitivity analysis. Plots show the fractional change in steady state values of growth (solid lines) and S:H biomass (dashed lines) in response to fractional changes in default parameter values (see Table 2 for default values). Parameters are grouped by the processes in which they are involved. This sensitivity analysis was conducted at conditions typical for coral reef environments: low DIN (1e-7 M) and intermediate light (15 mol photons m^-2^ d^-2^), with prey density set to zero. Sensitivity analyses conducted other environmental conditions are presented in Figs S3-S7.](../img/Fig4.png)

@@

![Light-driven seasonal dynamics of symbiont abundance and coral growth. Light input (upper panel) was designed as a sinusoidal curve with a period of one year, with maximum and minimum values of 44 and 20 mol photons m^-2^ d^-1^, corresponding to those measured by @Stimson:1997p3837. The dynamic behavior of symbiont to host biomass ratio (middle panel) and the specific growth rate of host biomass (lower panel) show seasonal oscillations that are greater in magnitude under high nutrients (15.14 $\mu$M N; dashed lines) relative to low nutrients (0.14 $\mu$M N; solid lines), consistent with the findings of @Stimson:1997p3837. Prey density was set at 1e-6 CmolX L^-1^.](../img/Fig5.png)

@@

![Coral bleaching as a switch from a nitrogen- to carbon-limited alternative stable state. This transition is demonstrated in response to gradually increasing external light (A), which causes an increase in production of ROS (B) that reduces the photosynthetic rate through photoinhibition (C). Decreasing photosynthesis moves the system from overall nitrogen-limitation toward carbon-limitation (D); when this threshold is crossed, the system rapidly becomes highly carbon-limited as photosynthesis becomes CO~2~-limited (E) and symbiont densities rapidly decline (F) into a bleached state. Substrate limitation coefficients were calculated using Eq. 23. All parameters were set at default values with external DIN=1e-7 mol N/L and prey density set to zero).](../img/Fig6.png)

@@

![Bleaching with interactive factors. Simulations of high light stress (sinusoid with maximum=48 mol ph m^-2^ d^-1^) under default environmental conditions (solid line; DIN=1e-7 mol N L^-1^; prey=2e-7 C-molX L^-1^), or with elevated feeding (dashed line; prey=1e-6 C-molX L^-1^), or elevated nutrients (dotted line; DIN=4e-6 mol N L^-1^). Initial symbiont biomass was set to the steady state for each set of starting conditions, with all other parameters at default values.](../img/Fig7.png)

@@

![Recovery from bleaching with varying N:C availability (A: N:C=0.100; B: N:C=0.175; C: N:C=0.220). Higher N:C ratios in a heterotrophic food source (effectively representing higher external DIN and/or lower feeding rates) cause a larger overshoot in the S:H biomass ratio, and prolong the duration of time until the system 'recovers' by re-establishing nitrogen-limitation. Substrate limitation coefficients were calculated using Eq. 23. All simulations were run with default parameters (except varying $n_{NX}$), with L=20 mol photons m^2^ d^-1^, DIN=1e-7 mol N L^-1^, prey=1e-7 CmolX L^-1^, and initial S:H biomass=0.001.](../img/Fig8.png)
